%Acknowledgments

\renewcommand{\baselinestretch}{2}
\small\normalsize
\hbox{\ }
 
\vspace{-.65in}

\begin{center}
\large{Acknowledgments} 
\end{center} 

\vspace{1ex}

I was very lucky to be surrounded by an amazing group of people throughout my PhD. None of this would have been possible without them and I feel infinitely lucky and grateful for that. 

I would like to start by thanking my advisor Dr. Ian Spielman for giving me the opportunity of joining his research group. Ian is one of the most brilliant scientists I have had the pleasure to meet and I was very fortunate to be able to work and learn with him. I am very grateful for all of his support, teaching, and for always being so patient and good-humored. Ian's unique advising style combining top-quality science with cheerful and laid back interactions makes the excitement for physics contagious!

I was exposed to a cold atoms lab for the first time when I spent a summer as an undergrad working in the cavity QED lab of Dr. Luis Orozco. I thank him for giving me this opportunity that would later define the course of my PhD. He is a great mentor who cares very deeply about his students. I am extremely grateful for his support, for all the delicious meals and bread, the concerts and plays, and the good stories. I also thank him for carefully reading my thesis and providing comments. 

I would like to thank Dr. Bill Phillips for always being warm and welcoming and for taking the time to participate in `paper torture' sessions with me. I learned a lot of physics (and history of physics) from these interactions. I would also like to thank him for being an endless source of chocolate. 

My research was performed at the RbLi lab at UMD, where I was lucky that my path crossed with many wonderful physicists and also now good friends. First I would like to thank  Daniel Campbell, Ryan Price and Andika Putra for welcoming me to RbLi and patiently teaching me about the lab. A very special thank you goes to our former postdoc Dimitri Trypogeorgos for taking me under his wing and becoming my lab brother. He is a brilliant scientist who really helped to shape my PhD. I am extremely grateful for all his physics and life teachings as well as his friendship. I would also like to thank and acknowledge my current lab mates. Qi-Yu Liang for her brutal honesty and for the physics discussions that tended to start with confusion and end with enlightenment. I also thank her for the rabbit pictures and videos. I admire her determination and her ability to stand up for what she believes in. Mingshu Zhao and Junheng Tao for their enthusiasm and diligently taking over the tasks of building a new BEC apparatus. I am also grateful for their encouraging words when I was writing this thesis and my last paper. They are very smart physicists and I am sure there will be great things to come in the new lab. 

Our lab was also very fortunate to have a couple brilliant visitors. I thank Nathan Lundblad for all his help when the lab was feeling a bit understaffed. I would also like to thank Russ `Mr. Broadband' Anderson for all the helpful experimental diagnostics gadgets, the very thorough help on calculations and comments on papers, and for always being in such a good mood. It was great fun to have him around. 

Other great visitors not directly working with me but that had an impact on my PhD include Marcell Gall, Pierre Dussarat, Max Schemmer, Daniel Ohl de Melo, and Pierre Bataille. I thank them for their friendship and for the good times we shared.  

I have been away from the lab many months now and instead spending a great deal of time in my office. I would like to thank my office mates for making it all more fun. Hector Soza Mart\'inez for being absolutely ridiculous. His sense of humor reminds me that we shouldn't take ourselves too seriously. Joao `the champ' Braz for being a great theory buddy. I really enjoyed our physics discussions and am thankful for all the thorough comments on the Rashba paper. I am also thankful that he made me leave my desk to get a beer when he thought I needed a break from working too hard. Dina Genkina for always listening and offering encouraging words, also for the constant flow of Kit Kats. 

I was very lucky to find a great community of Latin American physicists (and non-physicists) at UMD (thanks again Luis) who became some of my closest friends and kind of a family away from home. Having them around made everything more fun and helped me stay sane throughout my PhD. A big thank you to Paco, Ale, Pablo, Dalia, David, Israel, Kristel, Andres, Hector, Maggie, Carlos, Monica and Carla for their friendship. A special shout out to Dalia who has been like my academic sister ever since we both joined Luis' labs. Having a good friend like her sharing this PhD path really makes things better.  

Outside of the university, I was also very fortunate to run into a wonderful group of people, especially those affiliated to the extended International House of Physicists (IHoP) family: Juraj, Kanu, Yigit, Shanon, Anna, Marko, Aydn, Seyit, Daniel, Rory, Dina, Chris, Ankit, Ray. Having them around made life more fun and enjoyable! 

Besides the people I met in grad school, I had a network of good friends back home who have always been very supportive throughout the years. I thank Andrea, Adriana, Elsa, Guillermo, and Bernardo for always cheering me and making me feel like I was doing something cool. 

A common denominator in many aspects of my PhD has been my partner Paco. He has always been there for me through the best and worst times and I am very grateful for that. Paco is a very kind and caring person and also a great physicist. I also thank him for listening, making me laugh, and for all his help with physics and non-physics issues. I was very lucky to find a partner, friend, and colleague in him. 

I was deeply touched by the support I received for my thesis defense. Friends getting together to cook delicious tacos, surprise visits from Europe, family members visiting from various places, emails and text messages from friends and colleagues with good wishes. A special thanks to Kanu for also being my presentation coach. Her advise and `cups of tea' really helped me calm down when the nerves were kicking in hard. I don't want to sound repetitive but I am very lucky to be surrounded by such a nice community. 

Finally and most importantly, I would like to thank my family. My parents always made my education a priority and none of this would have been possible without their unconditional support. I also thank my brothers for their support and all the good (and bad) times that we have shared. They try hard to annoy me sometimes but I wouldn't change them for anyone else. 

